---
title: "Analysis for Hoecker and Higuera XXXX: "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results  = 'hide', message = FALSE, warning = FALSE)
```

## Setup

Load the requisite packages.
```{r}
library(tidyr) # For data manipulation
library(dplyr) # For data manipulation
library(zoo) # For rolling statistics 
library(Hmisc) # For correlation statistics
library(ggplot2) # For plotting
library(knitr) # For formatting this file 
```

Specify directory on your machine containing data (ends with 'nowitna/data') 
```{r}
dataDir <- "/Users/tylerhoecker/GitHub/nowitna/data/"
```

Specify lakes and time period of analysis.
```{r}
lakes <- c('BB15','DU15','MA15','NW15','SH15','TL15','UR15') 
studyPeriod <- c(1550,2015)
```

Load chronological data. Age-depth models were built in Bacon v2.2 (http://www.chrono.qub.ac.uk/blaauw/bacon.html), using the parameters described in the text (Table 2).  
```{r}
dates <- read.csv(paste0(dataDir,'/ageData_sample_inventory.csv'))

ageModel.df <- lapply(paste0(dataDir,'ageModel_',lakes,'.csv'), read.csv) %>% 
  `names<-` (lakes) %>%
  bind_rows(.id = 'lake')
```

Plot age-depth models.
```{r results = 'asis', echo = FALSE}
ggplot(ageModel.df, aes(x = depth)) + 
  geom_line(aes(y = median)) +
  geom_ribbon(aes(ymin = min, ymax = max), alpha = 0.3) +
  geom_point(data = dates, aes(x = TopDepth, y = CalibratedAge, shape = Type)) + #color = factor(Used)
  geom_errorbar(data = dates, 
                aes(x = TopDepth, ymin = CalAgeLCI, ymax = CalAgeUCI)) +
  geom_vline(data = dates, aes(xintercept = Conf), linetype = 2, color = 'grey20') +
  scale_shape_discrete('Sample\nType', breaks = c('C14','Pb210'), 
                       labels = c(as.expression(bquote(''^14~C)),
                                  as.expression(bquote(''^210~Pb)))) +
  scale_x_reverse() +
  scale_y_continuous(trans = 'reverse', sec.axis = sec_axis(~1950-., name = 'Median estimated age (Year CE)')) +
  coord_cartesian(ylim = c(-65,625)) +
  facet_wrap(~lake, ncol = 2, scales = 'free') +  #, scales = 'free'
  labs(x = 'Sample depth (cm)', y = 'Median estimated age (cal yr BP)') +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position = c(0.9, 0.1), # c(0,0) bottom left, c(1,1) top-right.
        legend.background = element_rect(fill = "#ffffffaa", colour = "black"))
```


Load charcoal data. These data have undergone peak analysis and interpolation in CharAnalysis (https://github.com/phiguera/CharAnalysis), using the parameters described in the text (Table 2).
```{r}
char.ldf <- lapply(paste0(dataDir,'charResults_',lakes,'.csv'),read.csv)   

char.df <- char.ldf %>%
  bind_rows() %>%
  mutate(ageCE = 1950 - ageTop,
         peakYr = ifelse(peaksFinal == 1,ageCE,NA),
         peakInsigYr = ifelse(peaksInsig == 1,ageCE,NA),
         peakYr1 = ifelse(peaks1 == 1,ageCE,NA),
         peakYr2 = ifelse(peaks2 == 1,ageCE,NA),
         peakYr3 = ifelse(peaks3 == 1,ageCE,NA)) %>%
  filter(ageCE >= studyPeriod[1]) %>%
  group_by(lake) %>%
         mutate(length = max(ageCE) - min(ageCE)) 
```

Load observed (historic) fire event data.
```{r}
observed.df <- read.csv(paste0(dataDir,'/observedfireData.csv'))
obs.1km <- filter(observed.df,distance == 1.0) %>%
  rename(obsCE = ageCE)
```

Load tree demography data.
```{r}
tree.df <- read.csv(paste0(dataDir,'treeData.csv'))
```

Load modeled and proxy climate data, bin into 5-year means.
```{r}
goa.df <- read.csv(paste0(dataDir,'climateData_GOA.csv'))
cru.df <- read.csv(paste0(dataDir,'climateData_CRU.csv'))
# Extract growing season temperatures from CRU data
cru.gs.df <- cru.df %>%
  filter(month %in% c(4:9)) %>%
  group_by(yearCE,variable) %>%
  summarise(mean = mean(value))
```


## Summarize fire event metrics 

Mean signal to noise indices of records during study period:
```{r results  = 'asis'}
kable(
  char.df %>% 
    summarise(meanSNI = mean(SNI, na.rm = T)),
  caption = 'Mean SNI'
)
```

Study-wide mean FRI (mean of all FRI pooled, rather than mean of means):
```{r results  = 'asis'}
FRI <- char.df %>%
  select(lake,peakYr,length) %>%
  filter(!is.na(peakYr)) %>%
  mutate(FRI =  (lead(peakYr, 1) - peakYr) * -1) 

kable(
  FRI %>%
    ungroup(lake) %>%
    summarise(min = min(FRI, na.rm = T),
              max = max(FRI, na.rm = T),
              mean = mean(FRI, na.rm = T),
              median = median(FRI, na.rm = T),
              sd = sd(FRI, na.rm = T)),
  caption = 'Study-wide mean FRI'
)
```

Individual mean FRI:
```{r results = 'asis'}
kable(
  FRI %>%
  group_by(lake) %>%
  summarise(nFires = n(),
            mFRI = mean(FRI, na.rm = T),
            sdFRI = sd(FRI, na.rm = T),
            length = length[1],
            FF = nFires/length * 100) ,
  caption = 'Mean FRI'
)
```

Calculate the difference (in years) between observed fire years within 1km of lakes and the most recent fire event detected in charcoal record
```{r results = 'asis'}
char.df <- char.df %>%
  group_by(lake) %>%
  right_join(.,obs.1km, by = 'lake')

kable(
  char.df %>% 
  group_by(lake) %>%
  summarise(observed = obsCE[1],
            lastPeak = max(peakYr, na.rm = T),
            difference = abs(obsCE[1] - max(peakYr, na.rm = T))),
  caption = 'Inferred vs. observed fire event timing.'
)
```

Plot lake-specific CHAR records (black line), inferred fire events (black dots), and theshold used to define events (grey line).
```{r results = 'asis', echo = FALSE}
ggplot(char.df, aes(x=ageCE, y=charAcc)) + theme_bw() +
  geom_step() +
  geom_line(aes(y = (treshPos+charBkg)), color = 'grey50') +
  geom_point(aes(x = peakYr, y = 0), shape = 21, size = 1.5, fill = 'black') +
  labs(y = bquote('Charcoal accumulation rate ('*pieces %.% cm^-2 %.% yr^-1*')'),
       x= 'Time (year CE)') +
  facet_wrap(~lake, scales = 'free', ncol = 2)
```



## Calculate proportion of sites burned through time, using 50-year windows in continuous 5-year time steps.
```{r}
window = 50
timeStep = 5

sitesByYear <- char.df %>%
  group_by(ageCE) %>%
  summarise(sites = n()) 

pctBurned <- char.df %>% 
  filter(!is.na(peakYr)) %>%
  group_by(peakYr) %>%
  summarise(n.burned = n()) %>%
  rename(ageCE = peakYr) %>%
  full_join(sitesByYear,.) %>%
  mutate(n.burned = ifelse(!is.na(n.burned),n.burned,0)) %>%
  # Window
  mutate(win.total = rollapply(sites, window/timeStep, fill= NA,
                               FUN = mean, na.rm =T)) %>%
  mutate(win.burn = rollsum(n.burned, window/timeStep, fill= NA)) %>%
  mutate(win.pct = win.burn/win.total*100) 


modernTime <- seq(1950,2015,1)
modernSites = data.frame('ageCE' = modernTime, 'sites' = 6)

pctModern <- observed.df %>%
  filter(distance == 1) %>%
  group_by(ageCE) %>%
  summarise(n.burned = n()) %>%
  full_join(modernSites,.) %>%
  mutate(n.burned = ifelse(!is.na(n.burned),n.burned,0)) %>%
  # Window
  mutate(win.total = rollapply(sites, window, fill= NA,
                               FUN = mean, na.rm =T)) %>%
  mutate(win.burn = rollsum(n.burned, window, fill= NA)) %>%
  mutate(win.pct = win.burn/win.total*100) 
```

Plot the result, a time series of FRI (grey squares), percent sites burned based on the charcoal record (black line), and percent sites burned based on observed fire data since 1950 (red line).
```{r echo = FALSE}
 ggplot(filter(pctBurned, ageCE>1550), aes(x = ageCE, y = win.pct)) +
  geom_step(data = pctModern, direction = 'vh', color = 'red3', size = 1) +
  geom_step(direction = 'vh') +
  geom_point(data = FRI, aes(x = peakYr, y = FRI/2), shape = 22, fill = 'grey50', size = 2) +
  scale_x_continuous(limits = studyPeriod, breaks = seq(studyPeriod[1],studyPeriod[2],50)) +
  scale_y_continuous(limits = c(0,100), sec.axis = sec_axis(trans = ~.*2,name = 'Fire return interval (yr)')) +
  labs(y = 'Percent of sites burned', x = 'Year CE') +
  theme_bw(base_size = 14)
```





## Build composite biomass burning recrod from individual records. 
Import raw charcoal count data.
```{r}
# Import selected regions / alaska 
dataframes <- lapply(paste0(dataDir,'charData_',lakes,'.csv'),read.csv) 
```
Derive and standardize charcoal accumulation rates (# cm2 yr-1, "CHAR").
```{r}
# Calculate and standardize CHAR 
for (i in 1:length(lakes)){
  # Calculate sediment accumulation rates (dataframes[[i]]$sedAcc) [cm/yr] 
  dataframes[[i]][,'sedAcc'] <- 
    (dataframes[[i]][,'cmTop']-dataframes[[i]][,'cmBot'])/
    (dataframes[[i]][,'ageTop']-dataframes[[i]][,'ageBot']) 
  # Calculate charcoal accumulation rates (dataframes[[i]]$char) [#/cm2/yr]
  dataframes[[i]][,'char'] <- 
    (dataframes[[i]][,'charCount']/dataframes[[i]][,'charVol'])*
    dataframes[[i]][,'sedAcc'] 
  # Log transform non-zero char
  nonZero <- which(dataframes[[i]][,'char'] > 0)
  transChar <- log(dataframes[[i]][nonZero,'char'])
  # Z-score transformed char
  zChar <- (transChar - mean(transChar, na.rm=T)) / sd(transChar, na.rm = T)
  # Exponentiate
  dataframes[[i]][nonZero,'char'] <- exp(zChar)
  # Average sample age (this added to work with RK scripts)
  dataframes[[i]][,'age'] <- 
    round(rowMeans(dataframes[[i]][,c('ageTop','ageBot')], na.rm = T))
  # Site/lake name
  dataframes[[i]][,'lake'] <- lakes[i]
}

# Turn multiple dataframes into one long data frame
dat <- do.call(rbind,lapply(dataframes,data.frame,stringsAsFactors=TRUE))
dat <- dat[,c('char','age','lake')]
```

Execute the method for building a composite biomass burning record used in Kelly et al. 2013 (10.1073/pnas.1305069110). The method estimates the parameters of a zero-inflated log-normal (ZIL) distribution of pooled charcoal counts in continuous moving windows of a user-defined width. In this analysis, 5-year and 50-year window widths are used (2.5 and 25 half-kernel widths, respectively).

This portion of the analysis is relegated to a separate R script for clarity of the workflow. Window widths and other parameters can be manipulated within the 'analysis_composite.r' script.

```{r message=FALSE}
source(file = file.path("analysis_composite.R"))
```

Plot the result, a composite record of biomass burning.
```{r results = 'asis', echo = FALSE}
ggplot(composite.df, aes(x=ageCE)) +
  geom_col(aes(y = highMean), fill = 'grey50', width = 5) +
  geom_ribbon(aes(ymin = lowCIlower, ymax = lowCIupper), alpha = 0.2) +
  geom_line(aes(y = lowMean)) +
  scale_x_continuous(limits = studyPeriod, 
                     breaks = seq(studyPeriod[1],studyPeriod[2],50)) +
  scale_y_continuous(breaks = c(0,1,2,3)) +
  labs(y = 'Biomass burning\n(standardized CHAR)', 
       x = 'Year CE') +
  coord_cartesian(ylim = c(0,3)) +
  theme_bw(base_size = 14) 
```


## Summarize tree demography data.
```{r}
tree.df %>%
  mutate(ageCE = plyr::round_any(pith,5)) %>%
  group_by(sp) %>%
  summarise(count = n()) 
```

Plot the result, a time series of estimated pith dates by species.
```{r results = 'asis', echo = FALSE}
speciesColors <- c("#E69F00", "#D55E00", "#4286f4","#009E73","#E69F00")

ggplot(tree.df, aes(x = pith)) + theme_bw(base_size = 12) +
  geom_histogram(stat = 'bin',
                 binwidth = 5,
                 position = 'stack', 
                 aes(fill = sp)) +
  scale_fill_manual(values = speciesColors) +
  scale_x_continuous(limits = studyPeriod, breaks = seq(studyPeriod[1],studyPeriod[2],50)) +
  scale_y_continuous(breaks = c(0,50,100)) +
  labs(x = 'Pith date / sample age (C.E.)', y = 'Count', fill = 'Species') 
```


## Compare proxies
Standardize and bin data to allow for direct comparison.
```{r}
# Modify tree dataframe 
tree.cor.df <- tree.df %>%
  mutate(ageCE = plyr::round_any(pith,5)) %>%
  group_by(sp, ageCE) %>%
  summarise(count = n()) %>%
  tidyr::spread(key = sp, value = count) %>%
  mutate(bepa = ifelse(is.na(bepa),0,bepa),
         lala = ifelse(is.na(lala),0,lala),
         pigl = ifelse(is.na(pigl),0,pigl),
         pima = ifelse(is.na(pima),0,pima),
         potr = ifelse(is.na(potr),0,potr)) %>%
  mutate(tree.count = rowSums(.[c('bepa','lala','pigl','pima','potr')])) 

# Use GOA mean from 1550-2010 to standardize both climate datasets. 
goa.1900 <- goa.df %>%
  filter(yearCE >= 1900) 
goa.mean <- mean(goa.1900$temp)

# Bin data into universal 5-year means
goa.df$yearBins <- cut(goa.df$yearCE,include.lowest = T,right = F,
                       breaks = seq(min(goa.df$yearCE),max(goa.df$yearCE),5), 
                       labels = seq(min(goa.df$yearCE),2005,5))
goa.binned <- goa.df %>%
  group_by(yearBins) %>%
  summarise(bin.temp = mean(temp)) %>%
  mutate(zscore = (bin.temp - goa.mean)) %>%
  mutate(sign = ifelse(zscore >= 0,'positive','negative'))


cru.gs.df$yearBins <- cut(cru.gs.df$yearCE,include.lowest = T,right = F,
                           breaks = seq(1900,2010,5), 
                           labels = seq(1900,2005,5))
cru.binned <- cru.gs.df %>%
  group_by(yearBins,variable) %>%
  summarise(bin.mean = mean(mean)) %>%
  group_by(variable) %>%
  mutate(zscore = (bin.mean - mean(bin.mean))) %>%
  mutate(sign = ifelse(zscore >= 0,'positive','negative'))

# Created combined dataframe of standardized, binned proxies for 1550-1895
goa.1550 <- goa.binned %>%
  mutate(ageCE = as.numeric(as.character(.$yearBins))) %>%
  select(-sign,-yearBins) %>%
  filter(ageCE < 1900)

combined.1550_1895.df <- composite.df %>%
  select(ageCE, lowMean, highMean) %>%
  inner_join(., pctBurned, by = 'ageCE') %>%
  inner_join(., goa.1550, by = 'ageCE') %>%
  left_join(., tree.cor.df, by = 'ageCE') %>%
  select(lowMean, highMean, win.pct, bin.temp, tree.count)

# Created combined dataframe of standardized, binned proxies for 1900-2005
goa.1900 <- goa.binned %>%
  mutate(ageCE = as.numeric(as.character(.$yearBins))) %>%
  select(-sign,-yearBins) 

cru.combined <- cru.binned %>%
  mutate(ageCE = as.numeric(as.character(yearBins))) %>%
  select(ageCE,variable,bin.mean) %>%
  spread(key = variable, value = bin.mean)

combined.1900_2010.df <- composite.df %>%
  inner_join(., pctBurned, by = 'ageCE') %>%
  inner_join(., cru.combined, by = 'ageCE') %>%
  inner_join(., goa.1900, by = 'ageCE') %>%
  left_join(., tree.cor.df, by = 'ageCE') %>%
  select(lowMean, highMean, win.pct, cru.precip = precip, cru.temp = temp, goa.temp = bin.temp, tree.count)
```

Plot standardized climate data.
```{r results = 'asis', echo = FALSE}
ggplot(goa.binned, 
         aes(x= as.numeric(as.character(yearBins)), y = zscore)) + theme_bw(base_size = 12) +
  geom_col(data = filter(cru.binned,variable == 'temp'), 
           aes(fill = sign, x= as.numeric(as.character(yearBins)), y = zscore)) +
  geom_step(aes(x= as.numeric(as.character(yearBins)) -1),direction = 'hv') +
  scale_fill_manual(values = c('negative' = 'dodgerblue4','positive'='red3')) +
  ylab('Growing season temp.\nanomaly (deg. C)') +
  scale_y_continuous(sec.axis = dup_axis(name = 'Modeled growing season temp.\nanomaly (deg.C)')) +
  scale_x_continuous(limits = studyPeriod, breaks = seq(studyPeriod[1],studyPeriod[2],50),
                     labels = c('',1600,'',1700,'',1800,'',1900,'',2000)) +
  xlab("Year CE") 
```




Calculate Spearman rank correlation among fire, tree, and climate records.
```{r}
corr.1550_1895.df <- rcorr(as.matrix(combined.1550_1895.df), type="pearson") 
corr.1900_2010.df <- rcorr(as.matrix(combined.1900_2010.df), type="pearson") 

```

Correlation coefficients and p-values for period 1550-1895 CE.
```{r results = 'asis', echo = FALSE}
kable(
  corr.1550_1895.df$r,
  caption = 'R'
)

kable(
  corr.1550_1895.df$P,
  caption = 'P-Value'
)
```


Correlation coefficients and p-values for period 1900-2010 CE.
```{r results = 'asis', echo = FALSE}
kable(
  corr.1900_2010.df$r,
  caption = 'R'
)

kable(
  corr.1900_2010.df$P,
  caption = 'P-Value'
)
```


Plot scatterplots of linear relationships between temperature and fire proxies for the two periods. 
```{r results = 'asis', echo = FALSE}
rp_label <- function(r,p){
  r.input = r; p.input = p
  text = substitute(italic(R)^2~"="~r.input*","~~italic(p)~"="~p.input,
                    list(r.input = format(r, digits = 2),
                         p.input = format(p, digits = 2)))
  as.character(as.expression(text))                 
}

r.text = corr.1550_1895.df$r['lowMean','bin.temp'] * corr.1550_1895.df$r['lowMean','bin.temp']
p.text = corr.1550_1895.df$P['lowMean','bin.temp']

ggplot(combined.1550_1895.df, aes(x = lowMean, y = bin.temp)) +
  geom_point(size = 3, color = 'black', fill = 'grey50', shape = 21) +
  theme_bw(base_size = 12) +
  geom_text(aes(x = .9, y = 7.8), label = rp_label(r.text,p.text), parse = T) +
  labs(x = '50-yr mean biomass burning', y = 'Reconstructed growing season temp.', title = '1550-1895') 

r.text = corr.1900_2010.df$r['lowMean','cru.temp'] * corr.1900_2010.df$r['lowMean','cru.temp']
p.text = corr.1900_2010.df$P['lowMean','cru.temp']

ggplot(combined.1900_2010.df, aes(x = lowMean, y = cru.temp)) +
  geom_point(size = 3, color = 'black', fill = 'grey50', shape = 21) +
  geom_smooth(method = 'lm', se = F, color = 'black', size = 1) +
  geom_text(aes(x = 1, y = 11.5), label = rp_label(r.text,p.text), parse = T) +
  theme_bw(base_size = 12) +
  labs(x = '50-yr mean biomass burning', y = 'Modeled growing season temp.',title = '1900-2015')
```









