---
title: "Analysis for Hoecker and Higuera XXXX: "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results  = 'hide', message = FALSE, warning = FALSE)
```

## Setup

Specify lakes and time period of analysis.
```{r}
lakes = c('BB15','DU15','MA15','NW15','SH15','TL15','UR15') 
studyPeriod <- c(1550,2015)
```

Load the requisite packages and data files for the analysis.
```{r}
library(dplyr) # For data manipulation
library(zoo) # For rolling statistics 
library(ggplot2) # For plotting
```

Load charcoal data (data should be stored in the same directory as 'analysis.Rmd'). These data have already undergone peak analysis and interpolation in CharAnalysis (https://github.com/phiguera/CharAnalysis). Raw charcoal count data and CharAnalysis input files available here: _____________________[Figshare Link?]
```{r}
char.ldf = lapply(paste0(lakes,'_charResults.csv'),read.csv)   

char.df <- char.ldf %>%
  bind_rows() %>%
  mutate(ageCE = 1950 - ageTop,
         peakYr = ifelse(peaksFinal == 1,ageCE,NA),
         peakInsigYr = ifelse(peaksInsig == 1,ageCE,NA),
         peakYr1 = ifelse(peaks1 == 1,ageCE,NA),
         peakYr2 = ifelse(peaks2 == 1,ageCE,NA),
         peakYr3 = ifelse(peaks3 == 1,ageCE,NA)) %>%
  filter(ageCE >= studyPeriod[1]) %>%
  group_by(lake) %>%
         mutate(length = max(ageCE) - min(ageCE)) 
```




You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
